---
title: "Final Report for Intro to Data Science"
output:
  html_notebook:
    toc: true
    toc_float: true
    number_sections: FALSE
---

# Introduction

Many scientists state that we are in the midst of a mass extinction, biodiversity has been lost at unprecedented rates compared to previous years, and with the increasing fears of climate change and humanity's effects on the environment these fears have only grown. Because of this I found particular interest in a data set that might serve to explain a pattern of what's going on, and especially one that might tell far lesser known stories with *significant* real-life applications.

The data set here originally finds it's origins as a the 2020-1 Extinct Species Report from [ICUN](#source_link), a group known for their work with the *Red List of Threatened Species* as well as being at the relative forefront of researches towards Red-List adjacent matters. While they do a number of reports this data is on the *Plantae* kingdom specifically, of which I felt more drawn towards due to being aquainted, and thus more confident in my abilities to have an in-depth analysis in. 

Now, without further into the hidden stories of this data set. 

# Prerequisites

Loading all required packages/data needed.
```{r}
#Import data set:

Extinct_Plants <- read.csv("https://github.com/LiminalAegis/final_project_SM2021/blob/master/data/Extinct_Plants.csv")

#Load packages:

library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
```
Data used for 'Extinct_Plants' can be found [here](#source_link): 


# Data Exploration

When initially presented with the data set I was pushed towards wanting a broad scope of understanding to later compress and narrow down into an individual level. As such, I wanted to get generally trends out of the way, to hopefully dig to the center of a hidden issue. 

General analysis:
```{r}
summary(Extinct_Plants)
```

Implemented the separate function in order to find Genus trends, being of importance to get a feel for a group of alike species to analyse later on.

```{r}
#Setting data up to notice genus trends
Extinct_Plants <- separate(Extinct_Plants, binomial_name, into =  c("Genus", "Species"), sep = " ")
```

And afterwards, came to one of my first big questions for exploration:
**Which Genus holds the largest number of extinct species?**

```{r}
Extinct_Plants %>%
  group_by(Genus) %>%
  summarize(extinction_total = n()) %>%
  arrange(desc(extinction_total))
```
And thus a breakthrough. The Cyanea genus is intriguing not only for being the top of the extinction totals, but being double the next highest Genus. Inhrently this leads to the next question:

**What circumstances would cause this?**

```{r}
Extinct_Plants %>%
  filter(Genus == "Cyanea") 
```
From this I knew ahead of time that continent climate seems to be consistent among all of Cyanea (as I get more into in [this section](#real_world)), so this left an analysis of the different threats each species seems to experience, as well as what the largest factor is for all of them. 

```{r}
Extinct_Plants %>%
  filter(Genus == "Cyanea")%>%
  select(threat_AA:threat_NA) %>%
  colSums(plants!= 0)
```
From this it's apparent that ISGD was the largest factor in influence but there were also other unanswered questions,
For one how consistently did all of the Genus experience threats?

```{r}

#Data transformation to get total threats each species faces
Threat_organizer <- Extinct_Plants %>% 
  filter(Genus == "Cyanea")%>%
  select(starts_with("threat"))
  threat_rows<- rowSums(Threat_organizer)
  
Cyanea_Extinct_Plants <- Extinct_Plants %>%
  filter(Genus == "Cyanea")
  
Cyanea_Extinct_Plants$threat_total = c(threat_rows)



Cyanea_Extinct_Plants %>%
  ggplot()+
  geom_bar(aes(x = threat_total , color = Species))+
   theme(legend.background = element_rect(fill="lightblue", 
                                  size=0.5, linetype="solid"))+
  theme(legend.position="bottom")+
  labs(
    title = "Threat amounts for Cyanea species",
    subtitle = "A glance into how consistently the Genus experiences ICUN defined threats",
    x = "Threat Count",
    y = "# of Species effected"
  )
  
  
```
Surprisingly it turned out having only a single threat was the norm, leaving questions about the consistency of these extinctions.
Creating the question of:
**What was the frequency of extinctions over time like?**

Which in turn lead to some data cleaning by making sure the "Before 1900" factor wouldn't lead to things being out of order due to years in this dataset being characters.

```{r}
Extinct_Plants$year_last_seen[Extinct_Plants$year_last_seen == "Before 1900"] <- 1900
```

```{r}
Extinct_Plants %>%
  ggplot()+
  geom_count(aes(y = year_last_seen, x = red_list_category))+
  labs(
    title = "Frequency of Extinction Types Per Year",
    subtitle = "As Observed Across Every Genus",
    y = "Years Since Last Seen",
    x = "Degree of extinction",
    fill = "Amount"
  )
```

```{r}
Cyanea_Extinct_Plants$year_last_seen[Cyanea_Extinct_Plants$year_last_seen == "Before 1900"] <- 1900

Cyanea_Extinct_Plants %>%
  ggplot()+
  geom_count(aes(y = year_last_seen, x = red_list_category))+
  labs(
    title = "Frequency of Extinction Types Per Year",
    subtitle = "As Observed in Cyanea",
    y = "Years Since Last Seen",
    x = "Degree of extinction",
    fill = "Amount"
  )
```
Upon seeing the results of the general trends I found interest in 2000-2002 as general trends indicated that being extinct only in the wild was generally the lesser of options, while the Cyanea trend demonstrating an especially interesting outlier with it's group given that in no other year was there a species explicitly extinct only in the wild. 

Thus, I sort the data down to learn about these problem outliers: 

```{r}
Cyanea_Extinct_Plants %>%
  filter(year_last_seen == "2000-2020") %>%
  select(-country,-continent, -group)
```

# Data analysis

Per my goal to reduce things as much as possible to get the story of the data at play here, I found myself reasearching for real-life connections to see how the data structure compares as well as reflecting notable circumstances. 

## Real world connections {.tabset .tabset-fade .tabset-pills #real_world}

### Cyanea Kuhihewa
### Cyanea Pinnatafinda 
### Cyanea Superba


# Conclusion

-yehaw im done

# Citations {#source_link}

* Link to dataset used (includes original source): <https://github.com/LiminalAegis/final_project_SM2021/tree/master/data>
* ICUN: <https://www.iucnredlist.org/>
* Christina Crooker <https://scholarspace.manoa.hawaii.edu/bitstream/10125/10404/uhm_ms_3912_r.pdf>
* FWS: <https://ecos.fws.gov/docs/five_year_review/doc1131.pdf>
* UNL: <https://digitalcommons.unl.edu/cgi/viewcontent.cgi?article=2548&context=icwdm_usdanwrc>
* DLNR: <https://dlnr.hawaii.gov/wildlife/files/2013/09/Fact-Sheet-Cyanea-pinnatifida.pdf> 
* DOI: <https://www.publicgardens.org/file/46700/download?token=4PbQJleN>

